Climate change projected to reduce European life expectancy \thanks{All data and code that supports these conclusions are available as supplementary materials.}

### 2018-04-05 MS-cclifeexpec.Rmd

---
output: 
  pdf_document:
    citation_package: natbib
    keep_tex: true
    fig_caption: true
    latex_engine: pdflatex
    template: LATEX/svm-latex-ms.tex
title: "Climate change projected to reduce European life expectancy"
thanks: "All data and code that supports these conclusions are available as supplementary materials."
author:
# - name: Mathew E. Hauer^1^*
#   affiliation: University of Georgia
# - name: Alexis R. Santos^2^
#   affiliation: Pennsylvania State University

@@

keywords: "Climate change, Life expectancy, Mortality, Demography, Excess mortality, Europe, Public health"
date: "`r format(Sys.time(), '%B %d, %Y')`"
geometry: margin=1in
#fontfamily: mathpazo
fontsize: 12pt
spacing: double
bibliography: LATEX/ccmort.bib
biblio-style: nature
fig_caption: true
# use apsr or nature
header-includes:
- \usepackage[all]{nowidow}
- \usepackage{rotating}
- \usepackage{fancyhdr}
- \usepackage[table]{xcolor}
- \usepackage{tabularx}
- \usepackage{makecell}
- \usepackage{xcolor}
- \pagestyle{fancy}
- \fancyhead{}
- \fancyhead[C]{Climate change life expectancy}

@@

###   Inputting user info for HMD
myusername <- getPass(msg = "HMD username: ", noblank = FALSE, forcemask = FALSE)
mypassword <- getPass(msg = "HMD password: ", noblank = FALSE, forcemask = FALSE)

@@

<!-- # Results -->
We find that climate change could alter life expectancy by `r round(mean(lt_lancet$DIF_MID),2)` years (-2.5 months (`r round(mean(lt_lancet$DIF_LOW),2)` - `r round(mean(lt_lancet$DIF_HIGH),2)` years)) in the average European country (**\autoref{figure1}**). This reduction is comparable to mortality due to influenza and pneumonia [@arias2013united] in the United States. Although the average European country could see a change of `r round(mean(lt_lancet$DIF_MID),2)` years, several countries are likely to experience considerably greater reductions in life expectancy. Luxembourg could see a reduction of up to `r round(lt_lancet$DIF_HIGH[which(lt_lancet$CNTRY == "LUX")],2)*-1` years and the medium-variant of climate hazards could cost the average Spaniard `r round(lt_lancet$DIF_MID[which(lt_lancet$CNTRY == "ESP")],2)*-1` years of life.

@@

>>>>

---
output: 
  pdf_document:
    citation_package: natbib
    keep_tex: true
    fig_caption: true
    latex_engine: xelatex
    template: LATEX/svm-latex-ms.tex
title: "Climate change projected to reduce European life expectancy"
thanks: "All data and code that supports these conclusions are available as supplementary materials."
author:
# - name: Mathew E. Hauer^1^*
#   affiliation: University of Georgia
# - name: Alexis R. Santos^2^
#   affiliation: Pennsylvania State University

@@

keywords: "Climate change, Life expectancy, Mortality, Demography, Excess mortality, Europe, Public health"
date: "`r format(Sys.time(), '%B %d, %Y')`"
geometry: margin=1in
#fontfamily: mathpazo
fontsize: 12pt
spacing: double
bibliography: LATEX/ccmort.bib
biblio-style: nature
# use apsr or nature
fig_caption: true
header-includes:
- \usepackage[all]{nowidow}
- \usepackage{rotating}
- \usepackage{fancyhdr}
- \usepackage[table]{xcolor}
- \usepackage{tabularx}
- \usepackage{makecell}
- \usepackage{xcolor}
- \pagestyle{fancy}
- \fancyhead{}
- \fancyhead[C]{Climate change life expectancy}

@@

###   Inputting user info for HMD

myusername <- getPass(msg = "HMD username: ", noblank = FALSE, forcemask = FALSE)
mypassword <- getPass(msg = "HMD password: ", noblank = FALSE, forcemask = FALSE)

@@

<!-- # Results -->
We find that climate change could alter life expectancy by `r round(mean(lt_lancet$DIF_MID),2)` years (-2.5 months (`r round(mean(lt_lancet$DIF_LOW),2)` to `r round(mean(lt_lancet$DIF_HIGH),2)` years)) in the average European country (**\autoref{figure1}**). This reduction is comparable to mortality due to influenza and pneumonia [@arias2013united] in the United States. Although the average European country could see a change of `r round(mean(lt_lancet$DIF_MID),2)` years, several countries are likely to experience considerably greater reductions in life expectancy. Luxembourg could see a reduction of up to `r round(lt_lancet$DIF_HIGH[which(lt_lancet$CNTRY == "LUX")],2)*-1` years and the medium-variant of climate hazards could cost the average Spaniard `r round(lt_lancet$DIF_MID[which(lt_lancet$CNTRY == "ESP")],2)*-1` years of life.

@@

<!-- # Results -->
We find that climate change could alter life expectancy by `r round(mean(lt_lancet$DIF_MID),2)` years (-2.5 months (`r round(mean(lt_lancet$DIF_LOW),2)` - `r round(mean(lt_lancet$DIF_HIGH),2)` years)) in the average European country (**Fig. 1**). Although the average European country could see a change of -0.22 years, several countries are likely to experience considerably greater reductions in life expectancy. Luxembourg could see a change of up to -2.2 years and the medium-variant of climate hazards could cost the average Spaniard 0.98 years of life.

The overall reduction in $e_0$ is comparable to mortality due to influenza and pneumonia [@arias2013united] in the United States. **Table 1** reports the ranking of Climate Change compared to the major causes of death (COD) in European countries using age-standardised death rates and the standard European population [@pace2013revision] for the year 2013. In three European countries (Spain, Luxembourg, Italy), Climate Change could become a top three COD. In Spain (high = `r round(standard$ASDR_high[which(standard$COUNTRY == "Spain")],2)`), Luxembourg (high = `r round(standard$ASDR_high[which(standard$COUNTRY == "Luxembourg")],2)`), France (high = `r round(standard$ASDR_high[which(standard$COUNTRY == "France")],2)`), and Italy (high = `r round(standard$ASDR_high[which(standard$COUNTRY == "Italy")],2)`) Climate Change could become the leading cause of death if Climate Change is more extreme than anticipated. Only all circulatory diseases combined and all cancers combined would cause more deaths.

```{r echo = FALSE, fig.cap=paste("Change in life expectancy at birth ($e_0$) due to business-as-usual climate change. We report changes in life expectancies due to climate change for twenty-eight European countries. The central values represent the ensemble median while the stems represent the upper and lower bounds of the inter-model climate variability.")}

Our results also suggest climate change mortality differentials are likely to unfold along highly uneven geographies. Whereas many Northern European countries could experience negligible impacts on life expectancy, five European countries could see life expectancy changes in excess of -0.5 years (Luxembourg, Spain, Italy, France, and Ireland). This group of countries could see life expectancy changes of more than -1.0 years if climate change hazards are more intense than anticipated. In some countries, climate change could thus become a bigger killer than trachea, bronchus, and lung malignant neoplasms (-0.85 years), acute myocardial infarction (-0.87 years), or all accident related mortality (-0.84 years) [@arias2013united] by the end of the century.

```{r echo = FALSE, warning= FALSE, message = FALSE, fig.cap=paste("Estimated reduction in life expectancy at birth ($e_0$) by the 2080s under the $MID$ scenario.")}

While Forzieri et al's [@forzieri2017increasing] results suggest that the greatest climate change related mortality will unfold along a north-south gradient, we find the greatest reduction in life expectancy unfolds along an east-west gradient (**Fig. 2**). The most westerly European countries tend to have the greatest reductions. Neighboring pairs of countries at similar latitudes could experience vastly different mortality regimes, but neighboring pairs of countries at similar longitudes seem to experience more similar mortality regimes from climate change. Emergent research in Atmospheric Rivers suggest Western Europe could be more susceptible to these types of events [@ramos2015daily]. Additionally, these results point to the importance of converting excess mortality into life expectancy to properly quantify the effects of climate change mortality.

<!-- # Discussion -->
In this article, we demonstrate the impact climate change could have on life expectancy at birth in twenty-eight European countries. Previous studies on climate change and excess mortality potentially misrepresent the impact climate change could have on human mortality. Contrary to excess mortality estimates, life expectancy is routinely used as a primary metric for communicating overall health outcomes and enjoys widespread use by major international organizations [@world2015world;@marmot2012building;@salomon2012healthy]. Life expectancy and its derivatives are the recommended metrics for population health [@parrish2010peer]. Additionally, it connects mortality estimates into intuitively understandable metrics, translating global estimates of mortality into individual outcomes. Our work reveals the extent to which climate change could reduce the average person's longevity by the end of the century, the first study of its kind, significantly expanding our understanding of climate change and public health; thus linking two of the major areas in current developmental and sustainability discussion at both national and international levels [@abel2016meeting].

### 2018-04-05 MS-cclifeexpec.tex

\author{\Large Mathew E. Hauer\textsuperscript{1}*\vspace{0.05in} \newline\normalsize\emph{University of Georgia}   \and \Large Alexis R. Santos\textsuperscript{2}\vspace{0.05in} \newline\normalsize\emph{Pennsylvania State University}  }

@@

\@ifpackageloaded{color}{
    \PassOptionsToPackage{usenames,dvipsnames}{color}
}{%
    \usepackage[usenames,dvipsnames]{color}
}
\makeatother
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={Mathew E. Hauer\textsuperscript{1}* (University of Georgia) and Alexis R. Santos\textsuperscript{2} (Pennsylvania State University)},
             pdfkeywords = {Climate change, Life expectancy, Mortality, Demography, Excess
mortality, Europe, Public health},  
            pdftitle={Climate change projected to reduce European life expectancy},
            colorlinks=true,
            citecolor=blue,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls

\usepackage[all]{nowidow}
\usepackage{rotating}
\usepackage{fancyhdr}
\usepackage[table]{xcolor}
\usepackage{tabularx}
\usepackage{makecell}
\usepackage{xcolor}
\pagestyle{fancy}
\fancyhead{}
\fancyhead[CE,CO]{Climate change life expectancy}

@@

{
   \vskip 13.5pt\relax \normalsize\fontsize{11}{12} 
\textbf{\authorfont Mathew E. Hauer\textsuperscript{1}*} \hskip 15pt \emph{\small University of Georgia}   \par \textbf{\authorfont Alexis R. Santos\textsuperscript{2}} \hskip 15pt \emph{\small Pennsylvania State University}

@@

\noindent Climate change related estimates of excess mortality clearly demonstrate
the dramatic impact on public health and human mortality from climate
change. However, life expectancy at birth is more easily communicated
and understood by the public, and controls for different geographic,
economic, and demographic groups. No studies have comprehensively
connected excess mortality due to climate change to anticipated
reductions in life expectancy. Without properly situating the potential
loss of life within the contexts of life expectancy, we risk
misrepresenting cost of climate change on mortality. In this paper, we
convert excess mortality estimates into potential reductions of life
expectancy at birth and report the cost of climate change on human
longevity. We find climate change could become the third largest life
expectancy reducer behind heart disease and cancer in some European
countries. Thus, the cost of inaction on climate change could approach
one year of life in some countries.

@@

\noindent \doublespacing \begin{itemize}
\tightlist
\item
  Corresponding author.
  \href{mailto:hauer@uga.edu}{\nolinkurl{hauer@uga.edu}}. p:
  706-542-9369.
\end{itemize}

\textsuperscript{1} Carl Vinson Institute of Government, University of
Georgia. 201 N. Milledge Ave. Athens, GA USA 30602.

\textsuperscript{2} Department of Sociology and Criminology,
Pennsylvania State University. 901 Oswald Tower, University Park, PA USA
16802.

\section{Introduction}\label{introduction}

Climate change's implications on humanity go far beyond estimates of
economic damages \citep{hsiang2017estimatin}, estimates of displacement
\citep{nicholls2011sea, hauer2016millions, hauer2017migration}, or human
conflict \citep{barnett2007climate} but have the potential to lead to
the loss of human life
\citep{forzieri2017increasing, pachauri2014climate}. As impact
quantification studies move further from the physical sciences of
climate change into the social sciences, properly quantifying and
conveying the impact of climate change on public health is of increasing
importance \citep{melillo2014climate, cloyd2016engagement}.

Scholars have long tried to estimate the mortality risks associated with
climate change and typically use excess or extra mortality
\citep{forzieri2017increasing, wilson2017climate, mcmichael2006climate, zanobetti2012summer}.
Although such estimates are useful, the excess mortality estimates are
rather sterile - one death is a tragedy, a million deaths a statistic -
and difficult to relate to on an individual level. Life expectancy at
birth, on the other hand, provides potent comparisons of mortality
vectors and converts excess mortality into an intuitively understood
metric, relatable to everyone. To date, no studies exist that
comprehensively examine the potential reductions of life expectancy due
to climate change. Without properly situating the potential loss of life
within the contexts of metrics such as life expectancy, we risk
underestimating the impact of climate change on human mortality.

Using published data on excess mortality, we connect climate change
excess mortality to life expectancy in a mortality model. By estimating
the increase in age-specific mortality rates associated with previously
published excess death estimates, we can assess how much climate change
could reduce the anticipated longevity of the average person in
twenty-eight European countries. This approach allows us to quantify the
impact of climate change on human longevity and answer the following
question: How much could climate change mortality reduce life expectancy
at birth? Our results can situate climate change mortality within the
broader context of human mortality and can be used to inform public
health interventions to prevent such futures.

We estimate changes in life expectancy by combining three primary
datasets: excess mortality data from Forzieri et al.
\citep{forzieri2017increasing}, cause-of-death distributions from the
Global Burden of Disease study (GBD) \citep{GBD, wang2012age}, and life
tables from the Human Mortality Database (HMD) \citep{HMD}. Forzieri et
al. \citep{forzieri2017increasing} focused on the projected excess
mortality of heatwaves, coldwaves, wildfires, droughts, river and
coastal floods, and windstorms under the business-as-usual greenhouse
gas emission scenario (SRES A1B). They modelled long-term, small-area
demographic changes in European countries that correspond to a
middle-of-the-road socioeconomic scenario (Shared-Socioeconomic Pathway
2 with medium fertility, medium mortality, medium migration, and the
Global Education Trend education scenario) and based their projected
mortality on extrapolations of an exhaustive examination of contemporary
disaster databases within spatially rectified, high-resolution gridded
datasets. While the disaster data are not standardized across country,
Forzeri et al. imputed data in incomplete time periods and countries.
This imputation could mask spatial variability at the sub-national
level, but here we use their country-aggregated results. Their results
represent business-as-usual climate change and human development without
incorporating potential adaptation. They found a potential 150,000+
climate change related fatalities per year by the mid 2080s with climate
change contributing to 90\% of mortality rise as opposed to population
changes. These data provide the anticipated fatalities but not the
effect these deaths might have on life expectancy.

To convert excess mortality to life expectancy, we use data gathered
from the HMD \citep{HMD} for corresponding European countries in
five-year life tables in conjunction with cause-specific mortality data
from the GBD \citep{GBD, wang2012age}. The HMD comprises only complete,
official vital event statistics and represents the most accurate and
complete compilation of mortality data available. We allocate excess
multi-hazard mortality for each country from supplementary table S8
\citep{forzieri2017increasing} based on the observed mortality schedule
for environmental heat and cold exposure deaths from the GBD for each
country in the previous decade.

We then recalculated age-specific mortality rates in five-year life
tables by adding the climate change mortality onto the underlying
mortality data and create four subsequent life tables (baseline, low,
mid, high) using standard life table techniques \citep{wunsch2013life}
based on the published confidence intervals from
\citep{forzieri2017increasing}. We used a variant of the cause-deleted
life tables approach
\citep{brand2005approximations, beltran2008integrated} to measure the
impact of excess mortality on life expectancy at birth or \(e_0\).
Changes in life expectancy are reported based on differentials from
\(e_0\) between the 2080s and the most recent complete life table in the
HMD. Complete life tables for all countries are available in the
Supplementary Dataset and all equations are located in the Supplementary
Information.

\section{Results}\label{results}

@@

\section{Discussion}\label{discussion}

@@

\caption{\textbf{Rankings of major causes of death in select European Countries based on a standardised death rate (per 100,000 inhabitants).} Climate Change COD is based on the 2080s while the other CODs are based on the 2010s.} 

@@

\vspace{1ex}
 \raggedright Data for other causes of death come from Eurostat (online data code: hlth_cd_asdr2). These results reflect the Standard European Population for 2013.

@@

Without adaptation measures, our results suggest climate change could
emerge as a significant new mortality vector and could pose a major
public health threat for some European countries by the end of the
century, echoing previous findings
\citep{forzieri2017increasing, patz2005impact}. Life expectancy has
steadily risen across the world for the last century
\citep{oeppen2002broken} and our results suggest that climate change
alone could spur a sharp reversal in these trends. We expect one
European countries to see life expectancy reductions more than one year
under the middle scenario (Luxembourg), but if climate change has a
greater impact on mortality than anticipated, five European countries
(Luxembourg, Spain, Italy, France, Ireland) could see life expectancy
reductions more than one year with Luxembourg experiencing a reduction
of more than two.

Reductions such as those should not be taken lightly. Many of the
children born today are likely to still be alive by the end of the
century and will be in the age groups most threatened by extreme heat
(65+) - the biggest mortality risk associated with climate change
\citep{keatinge2000heat}. If climate change unfolds as a more aggressive
mortality vector, only major cardiovascular diseases and malignant
neoplasms would contribute more to life expectancy reductions in
numerous European countries \citep{arias2013united}. This would make
climate change one of the most aggressive new mortality vectors to
emerge over the last quarter-century, representing a major threat to
public health in many parts of Europe.

@@

\section{Supplemenatary Information}\label{supplemenatary-information}

@@

To determine the differences between life expectancy compared to the
baseline, we simply subtract \(e_{x,i,s}\) (for scenario \(LOW\),
\(MID\), and \(HIGH\)) from \(e_{x,i,BASE}\) as described by
\citep{beltran2008integrated}.
\newpage
\singlespacing 
\bibliography{LATEX/ccmort}
\end{document}

>>>>

\author{}

@@

\@ifpackageloaded{color}{
    \PassOptionsToPackage{usenames,dvipsnames}{color}
}{%
    \usepackage[usenames,dvipsnames]{color}
}
\makeatother
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
             pdfkeywords = {Climate change, Life expectancy, Mortality, Demography, Excess
mortality, Europe, Public health},  
            pdftitle={Climate change projected to reduce European life expectancy},
            colorlinks=true,
            citecolor=blue,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls

\usepackage[all]{nowidow}
\usepackage{rotating}
\usepackage{fancyhdr}
\usepackage[table]{xcolor}
\usepackage{tabularx}
\usepackage{makecell}
\usepackage{xcolor}
\pagestyle{fancy}
\fancyhead{}
\fancyhead[C]{Climate change life expectancy}

@@

{
   \vskip 13.5pt\relax \normalsize\fontsize{11}{12}

@@

\noindent Climate change related estimates of excess mortality clearly demonstrate
the dramatic impact on public health and human mortality from climate
change
\citep{patz2005impact, aastrom2013attributing, forzieri2017increasing, staddon2014climate}.
However, life expectancy at birth is more easily communicated and
understood by the public, and controls for different geographic,
economic, and demographic groups \citep{parrish2010peer}. No studies
have comprehensively connected excess mortality due to climate change to
anticipated reductions in life expectancy. Without properly situating
the potential loss of life within the contexts of life expectancy, we
risk misrepresenting cost of climate change on mortality. In this paper,
we convert excess mortality estimates into potential reductions of life
expectancy at birth and report the cost of climate change on human
longevity. We find climate change could become the third largest life
expectancy reducer behind heart disease and cancer in some European
countries. Thus, the cost of inaction on climate change could approach
one year of life in some countries.

@@

\noindent \doublespacing \newpage

\section{Main Text}\label{main-text}

Climate change's implications on humanity go far beyond estimates of
economic damages \citep{hsiang2017estimating}, estimates of displacement
\citep{rigaud2018groundswell}, or human conflict
\citep{barnett2007climate} but have the potential to lead to the loss of
human life \citep{forzieri2017increasing, pachauri2014climate}. As
impact quantification studies move further from the physical sciences of
climate change into the social sciences, properly quantifying and
conveying the impact of climate change on public health is of increasing
importance \citep{melillo2014climate, cloyd2016engagement}.

Scholars have long estimated the mortality risks associated with climate
change and typically use excess or extra mortality
\citep{forzieri2017increasing, wilson2017climate, mcmichael2006climate, zanobetti2012summer}.
Although such estimates are useful, the excess mortality estimates are
rather sterile -- one death is a tragedy, a million deaths a statistic
-- and difficult to relate to on an individual level. Life expectancy at
birth (\(e_0\)), on the other hand, provides potent comparisons of
mortality vectors and converts excess mortality into an intuitively
understood metric, relatable to everyone. To our knowledge, no studies
exist that comprehensively examine the potential reductions of life
expectancy due to climate change. Without properly situating the
potential loss of life within the contexts of metrics such as life
expectancy, we risk underestimating the impact of climate change on
human mortality.

Using published data on excess mortality\citep{forzieri2017increasing},
we connect climate change excess mortality to life expectancy in a
mortality model. By estimating the increase in age-specific mortality
rates associated with previously published excess death estimates, we
can assess how much climate change could reduce the anticipated
longevity of the average person in twenty-eight European countries. This
approach allows us to quantify the impact of climate change on human
longevity and answer the following questions: What is the cost of
inaction on climate change on human longevity? and How does climate
change related mortality compare to other mortality vectors? Our results
can situate climate change mortality within the broader context of human
mortality and can be used to inform public health interventions to
prevent such futures.

We estimate changes in life expectancy by combining three primary
datasets: excess mortality data due to meterological hazards related to
climate change from Forzieri et al. \citep{forzieri2017increasing},
cause-of-death distributions from the Global Burden of Disease study
(GBD) \citep{GBD, wang2012age}, and life tables from the Human Mortality
Database (HMD) \citep{HMD}. Forzieri et al.
\citep{forzieri2017increasing} focused on the projected excess mortality
of heatwaves, coldwaves, wildfires, droughts, river and coastal floods,
and windstorms under the business-as-usual greenhouse gas emission
scenario (SRES A1B). They modelled long-term, small-area demographic
changes in European countries that correspond to a middle-of-the-road
socioeconomic scenario (SSP2 with medium fertility, medium mortality,
medium migration, and the Global Education Trend education scenario) and
based their projected mortality on extrapolations of an exhaustive
examination of contemporary disaster databases within spatially
rectified, high-resolution gridded datasets. While the disaster data are
not standardized across country, Forzeri et al. imputed data in
incomplete time periods and countries. This imputation could mask
spatial variability at the sub-national level, but here we use their
country-aggregated results. Their results represent business-as-usual
climate change and human development without incorporating potential
adaptation. They found a potential 150,000+ climate change related
fatalities per year by the mid 2080s with climate change contributing to
90\% of mortality rise as opposed to population changes. These data
provide the anticipated fatalities but not the effect these deaths might
have on life expectancy.

To convert excess mortality to life expectancy, we use data gathered
from the HMD \citep{HMD} for corresponding European countries in
five-year life tables in conjunction with cause-specific mortality data
from the GBD \citep{GBD, wang2012age}. The HMD comprises only complete,
official vital event statistics and represents the most accurate and
complete compilation of mortality data available. We allocate excess
multi-hazard mortality for each country from supplementary table S8 in
Forzieri et al. \citep{forzieri2017increasing} based on the observed
mortality schedule for environmental heat and cold exposure deaths from
the GBD for each country in the previous decade.

We then recalculated age-specific mortality rates in five-year life
tables by adding the climate change mortality onto the underlying
mortality data and create four subsequent life tables (\(baseline\),
\(low\), \(mid\), \(high\)) using standard life table techniques
\citep{wunsch2013life} based on the published confidence intervals from
\citep{forzieri2017increasing}. We used a variant of the cause-deleted
life tables approach
\citep{brand2005approximations, beltran2008integrated} to measure the
impact of excess mortality on life expectancy at birth or \(e_0\).
Changes in life expectancy are reported based on differentials from
\(e_0\) between the 2080s and the most recent complete life table in the
HMD. Complete life tables for all countries are available in the
Supplementary Dataset and all equations are located in the Supplementary
Information.

We find that climate change could alter life expectancy by -0.23 years
(-2.5 months (-0.1 to -0.39 years)) in the average European country
(\textbf{\autoref{figure1}}). This reduction is comparable to mortality
due to influenza and pneumonia \citep{arias2013united} in the United
States. Although the average European country could see a change of
-0.23 years, several countries are likely to experience considerably
greater reductions in life expectancy. Luxembourg could see a reduction
of up to 2.2 years and the medium-variant of climate hazards could cost
the average Spaniard 1.03 years of life.

\begin{figure}
\centering
\includegraphics{MS-cclifeexpec_files/figure-latex/figure1-1.pdf}
\caption{Change in life expectancy at birth (\(e_0\)) due to
business-as-usual climate change. We report changes in life expectancies
due to climate change for twenty-eight European countries. The central
values represent the ensemble median while the stems represent the upper
and lower bounds of the inter-model climate variability.\label{figure1}}
\end{figure}

Our results also suggest climate change mortality differentials are
likely to unfold along highly uneven geographies. Whereas many Northern
European countries could experience negligible impacts on life
expectancy, five European countries could see life expectancy changes in
excess of -0.5 years (Spain, Luxembourg, France, Italy, and Ireland).
This group of countries could see life expectancy changes of more than
-1.0 years if climate change hazards are more intense than anticipated.
In some countries, climate change could thus become a bigger killer than
trachea, bronchus, and lung malignant neoplasms (-0.85 years), acute
myocardial infarction (-0.87 years), or all accident related mortality
(-0.84 years) \citep{arias2013united} by the end of the century.

\begin{figure}
\centering
\includegraphics{MS-cclifeexpec_files/figure-latex/figure2-1.pdf}
\caption{Estimated reduction in life expectancy at birth (\(e_0\)) by
the 2080s under the \(MID\) scenario.\label{map}}
\end{figure}

\textbf{\autoref{Table1}} reports the rankings of Age Standardised Death
Rates (ASDR) for the leading causes of death in these European countries
(ASDR's for other CODs come from Eurostat (online data code:
hlth\_cd\_asdr2)) with our estimates of climate change included. In many
European countries, climate change could emerge as a top five killer.
Only combined CODs, such as all circulatory diseases or all cancers,
exhibit higher ASDRs if climatic hazards are more severe than
anticipated for Spain, Luxembourg, and France.

While Forzieri et al's \citep{forzieri2017increasing} results suggest
that the greatest climate change related mortality will unfold along a
north-south gradient, we find the greatest reduction in life expectancy
unfolds along an east-west gradient (\textbf{\autoref{map}}). The most
westerly European countries tend to have the greatest reductions.
Neighboring pairs of countries at similar latitudes could experience
vastly different mortality regimes, but neighboring pairs of countries
at similar longitudes seem to experience more similar mortality regimes
from climate change. Emergent research in Atmospheric Rivers suggest
Western Europe could be more susceptible to these types of events
\citep{ramos2015daily}. Additionally, these results point to the
importance of converting excess mortality into life expectancy to
properly quantify the effects of climate change mortality.

In this article, we demonstrate the impact climate change could have on
life expectancy at birth in twenty-eight European countries. Previous
studies on climate change and excess mortality potentially
miscommunicate the impact climate change could have on human mortality.
Contrary to excess mortality estimates, life expectancy is routinely
used as a primary metric for communicating overall health outcomes and
enjoys widespread use by major international organizations
\citep{world2015world, marmot2012building, salomon2012healthy}. Life
expectancy and its derivatives are the recommended metrics for
population health \citep{parrish2010peer}. Additionally, it connects
mortality estimates into intuitively understandable metrics, translating
global estimates of mortality into individual outcomes. Our work reveals
the extent to which climate change could reduce the average person's
longevity by the end of the century, the first study of its kind,
significantly expanding our understanding of climate change and public
health; thus linking two of the major areas in current developmental and
sustainability discussion at both national and international levels
\citep{abel2016meeting}.

@@

@@

\caption{\textbf{Rankings of major causes of death in select European Countries based on a standardised death rate (per 100,000 inhabitants).} Data for other causes of death comes from Eurostat.} \label{Table1}

@@

@@

Without adaptation measures, our results suggest climate change could
emerge as a significant new mortality vector and could pose a major
public health threat for some European countries by the end of the
century, echoing previous findings
\citep{forzieri2017increasing, patz2005impact}. Life expectancy has
steadily risen across the world for the last century
\citep{gerland2014world} and our results suggest that climate change
alone could spur a sharp reversal in these trends in some countries. We
expect two European countries to see life expectancy reductions more
than one year under the middle scenario (Luxembourg and Spain), but if
climate change has a greater impact on mortality than anticipated, five
European countries (Luxembourg, Spain, Italy, France, Ireland) could see
life expectancy reductions more than one year with Luxembourg
experiencing a reduction of more than two. These findings highlight the
hyperlocalized impacts of climate change
\citep{kendon2014heavier, rosenzweig2010cities, forzieri2017increasing}

Reductions such as those should not be taken lightly. Many of the
children born today are likely to still be alive by the end of the
century and will be in the age groups (aged 65+) most threatened by the
biggest mortality risk associated with climate change
\citep{keatinge2000heat} -- extreme heat. If climate change unfolds as a
more aggressive mortality vector, only all circulatory diseases combined
or all cancers combined would contribute more to mortality rates in
numerous European countries. This would make climate change one of the
most aggressive new mortality vectors to emerge over the last
quarter-century, representing a major threat to public health in many
parts of Europe.

@@

@@

To determine the differences between life expectancy compared to the
baseline, we simply subtract \(e_{x,i,s}\) (for scenario \(LOW\),
\(MID\), and \(HIGH\)) from \(e_{x,i,BASE}\) as described by
\citep{beltran2008integrated}.
\newpage
\singlespacing 
\bibliography{LATEX/ccmort.bib}
\end{document}
